On the Synchronization of Computational RFIDs

### 2018-05-03 response/response_letter.tex

in the name of all Authors---Dr. Przemys{\l}aw Pawe{\l}czak

\pagebreak

\begin{editor}
\referee{The authors need to revise the paper accordingly and prepare for the second review. The reviewers mainly concerns in the following aspects. \\
1. Clearly explain the motivation, describe clearly about the problem definition, and describe what new challenges address
in this paper. \\
2. Address the technical details mentioned by Reviewers 1 and 2. \\
3. More results to present mentioned by Reviewer 3.
}

\response{
	We believe that we revised our article in accordance with the responses of the reviewers. Please see our responses. 
}

@@

\end{editor}

\begin{responses}

\referee{This paper studies the problem of synchronization of computational RFID devices, which is a practically important issue. And the authors propose a lightweight synchronization method and implement it on the WISP tags. I still have the following concerns that require the authors to seriously consider. \\
\\	
In the Introduction part, the authors should clarify what type of synchronization they care. There are two types of synchronization of RFID communication: syn. between reader and tag; syn. among multiple tags. The COTS RFID devices exploit the EPC C1G2 standard, which uses framed slotted Aloha protocol to query tags. We found that the tag reading performance is very well. Hence, I think the synchronization among tags is not a big problem that hinder RFID communication. Hence, this study is not well motivated a least currently.}

\response{
	Please also see \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}.
}

@@

\referee{Also, there are another view to classify the synchronization problems: slot level synchronization (time frame is divided into slots) or bit level synchronization. Slot level synchronization has been studied for decades and is well addressed. The bit-level synchronization is not well investigated yet and attracts much attention from academic communities, e.g., the following three references. The authors should discuss and clarify whether they study the bit-level synchronization mentioned in these references. \\
**Top-k Queries for Categorized RFID Systems, IEEE/ACM Transactions on Networking (TON), 2017. \\
**Multi-category RFID Estimation, IEEE/ACM Transactions on Networking (TON), 2017. \\
**Completely Pinpointing the Missing RFID Tags in a Time-efficient Way, IEEE Transactions on Computers (TC), 2015.}

\response{
Thanks to these comments, we noticed that we did not highlight the main difference of our work from \emph{bit-level synchronization} in the previous submission. Moreover, we thank the reviewer for providing the studies we have not cited in the previous submission.\newline
As indicated by the reviewer, it is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal---i.e. bit-level synchronization. However, the objective of our work has the following fundamental differences as compared to these studies:
\begin{itemize}
	\item \textbf{Portability:} Bit-level synchronization relies on high-precision clock to obtain microsecond-precision timing measurements. Despite, in our work, we used we used the external 32\,KHz clock that can run in low-power standby mode and the \emph{higher-level} commands of the EPC-Gen 2 standard, rather than exploiting its low-level signal properties. Therefore, our approach is independent of how EPC Gen2 is implemented---providing a level of portability across different platforms. However, our synchronization algorithms can also be integrated to bit-level synchronization.
	\item \textbf{Explicit Clock:} Bit-level synchronization does not output an explicit clock value and it does not compensate for the clock drift. Rather, it uses offset compensation only and \emph{guard time} to reduce the effect of the clock drift. On the contrary, we provided time synchronization algorithms that compensate for the clock drift and provide an \emph{explicit} common time notion to CRFID applications.
	\item \textbf{Energy Budget:} Due to marginal energy budgets, we proposed synchronization algorithms that require minimum number of computation steps and memory space.
	\item \textbf{Power-Failures:} Providing synchronization service to the application layer requires to cope with power failures---which has not been investigated so far. To address this issue, we provided synchronization algorithms that can recover easily after power failures.
\end{itemize}

@@

\referee{What if WISP tags are simultaneously covered by multiple readers. In this case, the readers are not synchronized themselves and may have different time-stamps. Then, how does the proposed method work?}

\response{
As we indicated in the introduction of the article, there are synchronization techniques that provide synchronization among RFID readers. Therefore, if there are multiple readers that cover multiple WISP tags, first the readers should be synchronized with each other and then the tags will be synchronized to the RFID readers. We mentioned these issues in {\color{red} Section XXX} of our revised paper.
}

\referee{The novelty of this paper is relatively limited because the proposed event-based reader-tag synchronization method is originated from previous literature. Moreover, the main part of the theoretical analytics is also from previous work [24,31]. Although the authors claimed that they use a different clock drift model in this article, i.e. Brownian motion rather than uniform distribution, I still this significantly reduces the theoretical depth of this paper. The authors should show what NEW challenges beyond previous work are addressed by this paper. }

\response{
As we indicated in our previous responses (see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}), our article has the following novelties over the existing work in the literature:
\begin{itemize}
	\item Bit-level synchronization does not compensate for clock drift and it does not provide explicit time notion to the tags. In order to output a synchronized clock value, a time synchronization algorithm is required.
	\item Since the energy budget of the tags are limited, synchronization approaches that require marginal number of computation and memory. This is challenging---the main motivation for applying a PI-controller based  synchronization for the first time in this domain.
	\item Maintaining an explicit synchronized clock despite power failures has never been explored before in the field of CRFIDs. We also provided how our synchronization techniques overcome frequent power failures and provide a stable synchronized clock value.
	\item  We also provided an in-depth theoretical analysis of the proposed synchronization techniques. As indicate by the reviewer, the main part of the analysis is from previous work. However, the previous analysis that assumes uniform distribution for the clock drifts was not realistic---the reason we changed the analysis to use Brownian motion. This modification did not change the proof structure but it significantly changed the individual steps of the proof---that led to different stead-state error expressions.
\end{itemize}
}

@@

\referee{Some questions about the technical parts: \\
1): the authors claim that the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant clock drift in WISP. However, as far as I know, WISP uses a regulator to supply a constant 1.8V DC to power the WISP.}

\response{
	Fill
}

\referee{2): section 5.1, the authors observe a 1.89ms transmission delay. What does this delay account for? the in-air time should not be so long.
}

\response{
Since we used high-level properties of EPC Gen2 protocol, 
}

### 2018-05-04 response/response_letter.tex

\setlength{\parindent}{0pt}

\begin{document}

\pagestyle{myheadings}
\thispagestyle{empty}

\markright{\footnotesize TMC-2017-11-0695: {\sl On the Synchronization of Computational RFIDs}}

@@

\headsep 0.5cm

\bigskip\bigskip

%\noindent{Marwan Krunz, EIC, and Kevin Almeroth, Associate EIC \\
%	IEEE Transactions on Mobile Computing
%}

@@

\begin{flushright}
Dr. Przemys{\l}aw Pawe{\l}czak \\
Embedded Software Group \\ 
Delft University of Technology \\ 
Mekelweg 4, 2628 CD, Delft, The Netherlands \\
E-mail: p.pawelczak@tudelft.nl\\
Phone: +31 15 278 7491\\
\end{flushright}

\vspace*{2cm}

\today
\medskip

\textbf{TMC-2017-11-0695:} {\sl On the Synchronization of Computational RFIDs}

@@

\end{editor}

\begin{responses}

\referee{This paper studies the problem of synchronization of computational RFID devices, which is a practically important issue. And the authors propose a lightweight synchronization method and implement it on the WISP tags. I still have the following concerns that require the authors to seriously consider. \\
\\	
In the Introduction part, the authors should clarify what type of synchronization they care. There are two types of synchronization of RFID communication: syn. between reader and tag; syn. among multiple tags. The COTS RFID devices exploit the EPC C1G2 standard, which uses framed slotted Aloha protocol to query tags. We found that the tag reading performance is very well. Hence, I think the synchronization among tags is not a big problem that hinder RFID communication. Hence, this study is not well motivated a least currently.}

\response{
We thank the reviewer for pointing this issue that led us to emphasize our motivation in the revised article. As indicated by the reviewer, framed slotted Aloha protocol performs a notion of synchronization: by adjusting offset and introducing guard times, tags are ensured to communicate at allocated frame boundaries. However, 
\begin{itemize}
	\item this method does not output a synchronized clock value that can be accessed by the tags to timestamp and make inferences about the collected information;
	\item since a clock is not maintained, more sophisticated applications cannot be enabled.
\end{itemize}

In order to enable sensor network-like data collection applications, tags should maintain an explicit clock that represents synchronized notion of time so that they can  collect, timestamp and store data which will be be queried later by the readers. However, due to frequent power failures and marginal energy budgets of these devices, providing explicit clock synchronization and maintaining a clock value is challenging---please also see our responses \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}. We mentioned all these points in our revised article.
}

### 2018-05-04 response/response_letter.tex

\end{responses}

\begin{responses}

\referee{In this paper, the authors study the time synchronization problem in low-power backscatter systems, i.e., WISP. They point out the unique challenges in designing time synchronization algorithms for these energy-harvesting tags, and propose two synchronization protocols in response to these challenges. The proposed protocol is compatible with the default reader-to-tag communication protocol and consumes an ignorable amount of energy. The authors evaluate their proposed protocols through real-world experiments. \\
\\
+: an interesting problem which has not been explored before. \\
+: the proposed protocols are simple yet look effective. \\
+: evaluating the proposed protocols with real-world experiments; \\
\\		
-: the motivation behind is not well explained; \\
-: not sure the proposed protocol will work in large-scale deployment. \\
\\
My major concern about this paper is its motivation: I do agree with the authors that the time synchronization is important to the multi-hop backscatter networks, but I don’t think it is necessary for the single-hop network like the WISP networks. The ALOHA-type protocol has already synchronized the tags in nature. 
}

\response{ 
 The concern of this reviewer has been shared by also Reviewer 1. In our revised article, we provided a better motivation and highlighted the novelty of our work better.  \\
 In the previous responses---please also see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}---we mentioned that ALOHA-type protocol does not maintain an explicit clock value and does not implement a real synchronization algorithm. On the contrary, we provided novel synchronization mechanisms that enables tags to access synchronized time---this also enables sensor network-like data collection applications where tags collect and timestamp sensor readings that can be queried by the readers later. However, maintaining an explicit clock and ensuring synchronization is not trivial due to the marginal energy budgets of the tags and frequent power failures. These are the main challenges of our work and its novelty over existing studies. 
}

@@

\referee{I’m not sure synchronizing tags through the command initiated by the 
excitation generator (e.g., RFID reader) is feasible in real-world applications. This is because these energy-harvesting sensors can be charged within a very short range. Hence if the sensors are deployed sparsely, multiple excitation generators are needed, and it is thus important to synchronize these excitation generators as well. However, the authors did not consider this point.
}

\response{
We thank the reviewer for raising this issue. We already answered this concern in the previous responses---please also see \hyperref[R1:3]{[R1:3]}. We would like to mention that there are already prior work that provides synchronization among the readers. Therefore, we can also employ these strategies in order to create a common time notion among the readers. In order to provide an explicit synchronized clock value to the WISP tags, an explicit reader-tag synchronization algorithm is required---which is the focus of our work.
}

@@

\referee{Some questions about the technical parts: \\
1): the authors claim that the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant clock drift in WISP. However, as far as I know, WISP uses a regulator to supply a constant 1.8V DC to power the WISP.}

\response{
	Fill
}

\referee{2): section 5.1, the authors observe a 1.89ms transmission delay. What does this delay account for? the in-air time should not be so long.
}

\response{
In our article, we defined the \emph{transmission delay} as the time that passes between the start of the broadcast command from the reader and the receipt by the receiver node--- which is the major error source and composed of non-deterministic components. Since we used high-level properties of EPC Gen2 protocol, we could not use low-level timestamping mechanisms and our synchronization approaches suffer from big transmission delays.  The reason is that we do not know when the reader starts transmission after broadcast command is requested from the PC that controls the reader. During this process, the reader-side does not implement any timestamping mechanism and the delays are mainly dependent on the reader-side implementation of the EPC protocol.   
}

@@

\end{responses}

\begin{responses}

\referee{The authors have investigated the synchronization problem of RFID networks which consists of RFID readers and WISP tags. Several challenges of existing hardware platform and systems are identified and discussed. An event-based synchronization approach is proposed for solving the problem. A real demo system is implemented and used to evaluate the effectiveness of the proposed approach. The results show that the error of synchronization can be about 1ms and the energy overhead is about 1.85uJ. \\
\\
Generally, this paper is well written. It is a novel idea to improve the synchronization ability of RFID systems. Considering the emerging of internet of things, this work  is attractive.  The original version of this paper is published in a conference and this version includes sufficient additional contents for the journal. Therefore, I recommend this paper to be accepted after some revisions as follows:  \\
\\
1) It is better to simplify the descriptions of the challenges and existing works, especially some contents that are not directly related to the synchronization.}

\response{
	We would like to thank the reviewer for his/her encouraging comments.
}

### 2018-05-07 response/response_letter.tex

\response{
	We believe that we revised our article in accordance with the responses of the reviewers. Please refer to our respective responses below. 
}

\end{editor}

\begin{responses}

\referee{In the Introduction part, the authors should clarify what type of synchronization they care. There are two types of synchronization of RFID communication: syn. between reader and tag; syn. among multiple tags. The COTS RFID devices exploit the EPC C1G2 standard, which uses framed slotted Aloha protocol to query tags. We found that the tag reading performance is very well. Hence, I think the synchronization among tags is not a big problem that hinder RFID communication. Hence, this study is not well motivated a least currently.}

@@

\response{
We thank the reviewer for pointing this issue that led us to emphasize our motivation in the revised article. As indicated by the reviewer, framed slotted Aloha protocol performs a notion of synchronization: by adjusting offset and introducing guard times, tags are ensured to communicate at allocated frame boundaries. However, 
\begin{itemize}
	\item this method does not output a synchronized clock value that can be accessed by the tags to timestamp and make inferences about the collected information;
	\item since a clock is not maintained, more sophisticated applications cannot be enabled.
\end{itemize}

In order to enable sensor network-like data collection applications, tags should maintain an explicit clock that represents synchronized notion of time so that they can  collect, timestamp and store data which will be be queried later by the readers. However, due to frequent power failures and marginal energy budgets of these devices, providing explicit clock synchronization and maintaining a clock value is challenging---please also see our responses \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}. We mentioned all these points in our revised article.
}

\referee{Also, there are another view to classify the synchronization problems: slot level synchronization (time frame is divided into slots) or bit level synchronization. Slot level synchronization has been studied for decades and is well addressed. The bit-level synchronization is not well investigated yet and attracts much attention from academic communities, e.g., the following three references. The authors should discuss and clarify whether they study the bit-level synchronization mentioned in these references.
\begin{itemize}
\item \textbf{Top-k Queries for Categorized RFID Systems}, IEEE/ACM Transactions on Networking, 2017;
\item \textbf{Multi-category RFID Estimation}, IEEE/ACM Transactions on Networking, 2017;
\item \textbf{Completely Pinpointing the Missing RFID Tags in a Time-efficient Way}, IEEE Transactions on Computers, 2015.
\end{itemize}}

\response{
Thanks to these comments, we noticed that we did not highlight the main difference of our work from \emph{bit-level synchronization} in the previous submission. Moreover, we thank the reviewer for providing the studies we have not cited in the previous submission---we have added these papers to our related work list.

As indicated by the reviewer, it is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal---i.e. bit-level synchronization. However, the objective of our work has the following fundamental differences as compared to these studies:
\begin{itemize}
	\item \textbf{Portability:} Bit-level synchronization relies on high-precision clock to obtain microsecond-precision timing measurements. Despite, in our work, we used we used the external 32\,KHz clock that can run in low-power standby mode and the \emph{higher-level} commands of the EPC-Gen 2 standard, rather than exploiting its low-level signal properties. Therefore, our approach is independent of how EPC Gen2 is implemented---providing a level of portability across different platforms. However, our synchronization algorithms can also be integrated to bit-level synchronization.
	\item \textbf{Explicit Clock:} Bit-level synchronization does not output an explicit clock value and it does not compensate for the clock drift. Rather, it uses offset compensation only and \emph{guard time} to reduce the effect of the clock drift. On the contrary, we provided time synchronization algorithms that compensate for the clock drift and provide an \emph{explicit} common time notion to CRFID applications.
	\item \textbf{Energy Budget:} Due to marginal energy budgets, we proposed synchronization algorithms that require minimum number of computation steps and memory space.
	\item \textbf{Power-Failures:} Providing synchronization service to the application layer requires to cope with power failures---which has not been investigated so far. To address this issue, we provided synchronization algorithms that can recover easily after power failures.
\end{itemize}

### 2018-05-12 TMC-2017-11-0695-main.tex
finished

As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}. However, recent advancements in the wireless communication enable tag-to-tag communication, e.g.~\cite{liu2013ambient}. We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts. In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences at the CRFID tag side. Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.

\subsection{Differences From Battery-Powered WSNs }
\label{sec:Motivation}

However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service. The reason is mainly twofold:

@@

The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered CRFID systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices. In particular, the main contributions and findings of 
this article are:
%
\begin{itemize}
	\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems. Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides synchronization of individual tags in the communication range of an RFID reader. 
	\item By addressing the aforementioned challenges, we design and implement \emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015,yildirim2017adaptive}. WISP-Sync (i) 
	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
	%
	\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
	%
	\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
\end{itemize}
%
We note that WISP-side implementations throughout the article are done using C and assembly language. All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\href{https://goo.gl/RuWIZ2}{https://goo.gl/RuWIZ2}.

@@

\subsubsection{Wireless Communication}

@@

The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information. Commonly, \emph{least-squares regression} is employed by several practical synchronization protocols to adjust the offset and the frequency of the software clocks ~\cite{Maroti2004,Lenzen:2015,TATS:2016}. On the other hand \emph{iterative} computation methods, e.g. in ~\cite{Yildirim:AVTS:2014,pi2015,yildirim2017adaptive,Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round. With iterative methods, sensor nodes establish synchronization after a finite number of rounds in a lightweight manner in terms of computation and memory requirements. There are also \emph{fully distributed} approaches in which sensor nodes interact only with and synchronize to their direct neighbors in a peer-to-peer fashion without the requirement of a dedicated reference node. In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}. Since the prominent factor affecting the clock frequencies in WSNs is the temperature~\cite{schmid2010case}, there are also temperature compensated time synchronization techniques, e.g. \cite{Schmid:2009}. However, such techniques require a calibration phase in order to build a temperature--clock frequency table.
\TODO{Cite **Top-k Queries for Categorized RFID Systems, IEEE/ACM Transactions on Networking (TON), 2017. \\
	**Multi-category RFID Estimation, IEEE/ACM Transactions on Networking (TON), 2017. \\
	**Completely Pinpointing the Missing RFID Tags in a Time-efficient Way, IEEE Transactions on Computers (TC), 2015.}

@@

\subsubsection{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication. Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}. Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is impossible.

@@

\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term. On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift. Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.

\subsubsection{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g. the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up. However, saving computational state is also an energy consuming task~\cite{blisp}.

\subsubsection{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p. 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}. Due to the intermittent power, lightweight methods in terms of computation and memory are required for the synchronization. Since methods like least-squares regression are computationally heavy and require considerable amount of memory~\cite{pi2015} they should be avoided.

@@

\subsubsection{The Lack of broadcast primitive} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader. Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.

@@

Even though the event-based approach is relatively lightweight and adaptive as compared to the sender-receiver synchronization, it has some limitations. First, the tag is unable to obtain an \emph{explicit reference clock value}. The only variable it tunes is the $\hat{f}_w^r$ that represents the relative frequency of the software clock with respect to the clock of the reader. Therefore, we require additional steps that will allow tags to obtain the actual reference clock value. However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag. Second, WISP-Sync requires knowledge about $\tau$, which can be RFID reader dependent and requires a sniffer setup described in Section~\ref{sec:sniffer} to measure. This limitation can be addressed by employing an iterative estimation as an initial step to the WISP-Sync: the tag can start with an initial estimated $\tau$ value, and improve its estimation as it receives successive packets from the reader. After some iterations, the tag can have a rough estimate $\hat{\tau}$ that can be used on behalf of $\tau$. However, estimation errors might also perturb the synchronization accuracy. Last, the steps in Fig.~\ref{fig:event_based} present the communication scenario between one WISP and the reader. To allow other tags sniff this communication and synchronize, a broadcast primitive is crucial---in the next section we show how to overcome this limitation.

>>>>

As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}. \add{Therefore, CRFID tags cannot
operate in a multi-hop manner.} However, recent advancements in the wireless communication enabled tag-to-tag communication, e.g.~\cite{liu2013ambient}.  We anticipate that \add{this communication capability will lead up multi-hop CRFID networks that} will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.

\add{Since we are not aware of any network infrastructure to demonstrate neither multi-hop communication nor multi-hop synchronization, our focus in this article is the networks where several readers cover a tag population simultaneously. A representative greenhouse monitoring application where several tags are powered by two readers is depicted in Figure~\ref{fig:motivation}. In this application, the tags sense the environment and timestamp their temperature readings. When the readers request their temperature readings, the tags transmit their timestamps together with the collected temperature data. In order to make inferences on the collected temperature readings, timestamps should be consistent so that the readers can provide a temporal ordering of the readings. Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization. In order to achieve synchronization, first readers should acquire the global time, .e.g. by accessing time servers using the Internet; then individual tags should be synchronized to the individual readers.}

%We conjecture that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences at the CRFID tag side. Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.

\begin{figure}
	\centering
	\includegraphics[width=0.7\columnwidth]{figures/scenario.pdf}
	\caption{\label{fig:motivation}\color{blue}A representative greenhouse monitoring application using CRFIDs. Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings.}
\end{figure}

\subsection{Differences From Battery-Powered WSNs }
\label{sec:Motivation}
\add{The EPC Gen2 uses slotted Aloha protocol that performs a form of synchronization between the reader and the tags: by adjusting clock offset differences and introducing guard times, tags are ensured to communicate at the allocated slot boundaries. However, this method does not compensate for clock drifts and output a synchronized clock value that can be accessed by the tags to timestamp (and make inference about) the collected information, e.g. from the sensors attached to tags. In consequence, applications requiring time notion cannot be enabled, since a clock should be maintained at each tag---making synchronization service mandatory. However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service. The reason is mainly twofold:}

@@

The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered CRFID systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices. In particular, the main contributions and findings of 
this article are:
%
\begin{itemize}
	\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems. Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides synchronization of individual tags in the communication range of an RFID reader. 
	\item By addressing the aforementioned challenges, we design and implement \emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015,yildirim2017adaptive}. WISP-Sync (i) 
	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
	\item \add{Our time synchronization algorithms compensate for the clock drift and provide access to an \emph{explicit clock} for CRFID applications. Moreover, our implementations are independent of how EPC Gen2 is implemented ---providing a level of portability across different platforms.}
	%
	\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
	%
	\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
\end{itemize}
%
We note that WISP-side implementations throughout the article are done using C and assembly language. All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\href{https://goo.gl/RuWIZ2}{https://goo.gl/RuWIZ2}.

@@

\subsubsection{Synchronization Error}

@@

The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information. Commonly, \emph{least-squares regression} is employed by several practical synchronization protocols to adjust the offset and the frequency of the software clocks ~\cite{Maroti2004,Lenzen:2015,TATS:2016}. On the other hand \emph{iterative} computation methods, e.g. in ~\cite{Yildirim:AVTS:2014,pi2015,yildirim2017adaptive,Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round. With iterative methods, sensor nodes establish synchronization after a finite number of rounds in a lightweight manner in terms of computation and memory requirements. There are also \emph{fully distributed} approaches in which sensor nodes interact only with and synchronize to their direct neighbors in a peer-to-peer fashion without the requirement of a dedicated reference node. In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}. Since the prominent factor affecting the clock frequencies in WSNs is the temperature~\cite{schmid2010case}, there are also temperature compensated time synchronization techniques, e.g. \cite{Schmid:2009}. However, such techniques require a calibration phase in order to build a temperature--clock frequency table.

@@

\subsubsection{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication. Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}. Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is not feasible.

@@

\add{\subsubsection{Slot and Bit-level Synchronization}
EPC Gen2 requires slot-level synchronization to implement slotted Aloha protocol. This is achieved by adjusting clock offsets upon reception from the reader and introducing guard times to ensure that tags communicate at the
allocated frame boundaries. It is also possible to exploit modulation
properties of signals from the reader to the tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal, i.e. bit-level synchronization~\cite{liu_tcomp_2015,liu_ton_2017,liu2_ton_2017}. However, neither slot nor bit level synchronization compensate for clock drifts and output a synchronized clock value that can be accessed by the tags. 
}

\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term. On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift. \add{Even though embedded platforms are equipped with voltage regulators, the output of these regulators also exhibit instability and inaccuracy with respect to the input voltage; i.e. they are not perfect and they have an input voltage transient response. Therefore, they have an output
voltage variation with respect to the input voltage---still leading to clock frequency instabilities.} Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.

\subsubsection{Frequent power failures} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g. the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up. However, saving computational state is also an energy consuming task~\cite{blisp}.

\subsubsection{Energy and resource constraints} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p. 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}. Due to the intermittent power, lightweight methods in terms of computation and memory are required for the synchronization. Since methods like least-squares regression are computationally heavy and require considerable amount of memory~\cite{pi2015} they should be avoided.

@@

\subsubsection{The Lack of broadcast primitive} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader. 
%Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.

@@

Even though the event-based approach is relatively lightweight and adaptive as compared to the sender-receiver synchronization, it has some limitations. First, the tag is unable to obtain an \emph{explicit reference clock value}. The only variable it tunes is the $\hat{f}_w^r$ that represents the relative frequency of the software clock with respect to the clock of the reader. Therefore, we require additional steps that will allow tags to obtain the actual reference clock value. \add{Moreover, if the tags are covered by different readers, the first step should be first synchronizing the readers to the real-time. e.g. by allowing them to access time servers using internet.}
%However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag.
Second, WISP-Sync requires knowledge about $\tau$, which can be RFID reader dependent and requires a sniffer setup described in Section~\ref{sec:sniffer} to measure. This limitation can be addressed by employing an iterative estimation as an initial step to the WISP-Sync: the tag can start with an initial estimated $\tau$ value, and improve its estimation as it receives successive packets from the reader. After some iterations, the tag can have a rough estimate $\hat{\tau}$ that can be used on behalf of $\tau$. However, estimation errors might also perturb the synchronization accuracy. Last, the steps in Fig.~\ref{fig:event_based} present the communication scenario between one WISP and the reader. To allow other tags sniff this communication and synchronize, a broadcast primitive is crucial---in the next section we show how to overcome this limitation.
